---
title: "Analysyis of NPX format Olink data"
author: "Geoff Winsor"
date: "2024-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Gene set enrichment analysis (GSEA) of Olink data using OlinkAnalyze and clusterProfiler


```{r}
library(OlinkAnalyze)   
library(dplyr)
library(ggplot2)
library(stringr)
library(clusterProfiler)

#install.packages(
#  "msigdbdf",
#  repos = c("https://igordot.r-universe.dev", "https://cloud.r-project.org")
#)


library(msigdbr)

```



### Read NPX data (read_NPX)
```{r}
data <- read_NPX("data/2127B Exported NXP Values.xlsx")
```




### Input the exposure data file used for the other anlayses but keep only the variables of interest
```{r}
alldata <- read.csv("data/aim2_exwas_data_2024-June-11.csv")


df = read.csv('data/CHILD_ML_Study_variables_selected_and_transformed_A.csv')
df <- df %>% select(c('Subject_Number','PSS_18week_High_Stress','PSS_18week_Moderate_Stress','PSS_18week_Low_Stress','PSS_18week_Missing'))

df$subject_id <- df$Subject_Number
df$Subject_Number<-NULL
df$PSS_18week_High_Stress<-as.numeric(as.logical(df$PSS_18week_High_Stress))
df$PSS_18week_Moderate_Stress<-as.numeric(as.logical(df$PSS_18week_Moderate_Stress))
df$PSS_18week_Low_Stress<-as.numeric(as.logical(df$PSS_18week_Low_Stress))
df$PSS_18week_Missing<-as.numeric(as.logical(df$PSS_18week_Missing))
df <- df %>% filter(PSS_18week_Missing==0)

alldata<- merge(alldata, df, by.x = "subject_id", by.y = "subject_id", all.x = TRUE, all.y = FALSE)



alldata$adrenic_acid<-alldata$log_C22.4n6_3m
alldata <- alldata %>% select(c('subject_id','weekly_furniture_floor_dusting_polish_use_mother_prenatal','atopic_dermatitis_status_1y','atopy_binary_1y','adrenic_acid','monthly_furniture_floor_dusting_polish_use_prenatal','daily_hand_sanitizer_used_prenatal','prenatal_antibacterial','asthma_5y','recurrent_wheeze_binary_1y','clean_home_more_than_4_times_per_month_3m','log_dehp_bc_rec_dust_3m','family_asthma_history','daily_hand_sanitizer_used_5Y','vitamins_supplements_3m','weekly_chem_hand_cleaner_use_mother_prenatal','weekly_oven_cleaner_use_mother_prenatal','frequent_baby_lotion_used_3m','weekly_or_more_bleach_use_prenatal','is_male','is_caucasian','BF_3m','vaginal_birth','PSS_18week_High_Stress','PSS_18week_Moderate_Stress','PSS_18week_Low_Stress'))


vitamins_d_use_df = read.csv("data/vitamin_D_use_by_infant_all_sources_3m.csv")
alldata<- merge(alldata, vitamins_d_use_df, by.x = "subject_id", by.y = "subject_id", all.x = TRUE, all.y = FALSE)
```


### Generate the variable representing exposure to house dust DEHP in the upper quartile. 
```{r}
# Calculate the upper quartile (75th percentile) of the variable
upper_quartile <- quantile(alldata$log_dehp_bc_rec_dust_3m, 0.75, na.rm = TRUE)
alldata$upper_quartile_indicator <- alldata$log_dehp_bc_rec_dust_3m > upper_quartile
alldata$upper_quartile_indicator <- as.integer(alldata$upper_quartile_indicator)

```


### Generate the variable representing exposure to adrenic acid in the upper quartile. 
```{r}
# Calculate the upper quartile (75th percentile) of the variable
upper_quartile_adrenic_acid <- quantile(alldata$adrenic_acid, 0.75, na.rm = TRUE)
alldata$upper_quartile_adrenic_acid <- alldata$adrenic_acid > upper_quartile_adrenic_acid
alldata$upper_quartile_adrenic_acid <- as.integer(alldata$upper_quartile_adrenic_acid)

```



### Set the binary variables as factors
```{r}

alldata$vitamins_supplements_3m<-as.factor(alldata$vitamins_supplements_3m)
alldata$daily_hand_sanitizer_used_5Y<-as.factor(alldata$daily_hand_sanitizer_used_5Y)
alldata$weekly_chem_hand_cleaner_use_mother_prenatal<-as.factor(alldata$weekly_chem_hand_cleaner_use_mother_prenatal)
alldata$monthly_furniture_floor_dusting_polish_use_prenatal<-as.factor(alldata$monthly_furniture_floor_dusting_polish_use_prenatal)
alldata$weekly_oven_cleaner_use_mother_prenatal<-as.factor(alldata$weekly_oven_cleaner_use_mother_prenatal)

alldata$PSS_18week_High_Stress<-as.factor(alldata$PSS_18week_High_Stress)
alldata$PSS_18week_Moderate_Stress<-as.factor(alldata$PSS_18week_Moderate_Stress)
alldata$PSS_18week_Low_Stress<-as.factor(alldata$PSS_18week_Low_Stress)

alldata$upper_quartile_adrenic_acid<-as.factor(alldata$upper_quartile_adrenic_acid)
alldata$frequent_baby_lotion_used_3m<-as.factor(alldata$frequent_baby_lotion_used_3m)
alldata$is_male<-as.factor(alldata$is_male)
alldata$family_asthma_history<-as.factor(alldata$family_asthma_history)
alldata$BF_3m<-as.factor(alldata$BF_3m)
alldata$prenatal_antibacterial<-as.factor(alldata$prenatal_antibacterial)
alldata$vaginal_birth<-as.factor(alldata$vaginal_birth)
alldata$weekly_or_more_bleach_use_prenatal<-as.factor(alldata$weekly_or_more_bleach_use_prenatal)
alldata$daily_hand_sanitizer_used_prenatal<-as.factor(alldata$daily_hand_sanitizer_used_prenatal)
alldata$upper_quartile_indicatoxfr<-as.factor(alldata$upper_quartile_indicator)
alldata$asthma_5y<-as.factor(alldata$asthma_5y)
alldata$weekly_furniture_floor_dusting_polish_use_mother_prenatal<-as.factor(alldata$weekly_furniture_floor_dusting_polish_use_mother_prenatal)
alldata$vitamin_D_taken_by_infant_3m<-as.factor(alldata$vitamin_D_taken_by_infant_3m)
```


### Left join on the eligible subjects data to restrict the analysis to consented participants only (and get biological sex and study centre)
```{r}
eligible_subjects <- read.csv("data/eligible_subjects.csv")
eligible_subjects <- eligible_subjects %>% select(c('subject_id','gender','study_centre'))
eligible_subjects$gender<-as.factor(eligible_subjects$gender)
eligible_subjects$study_centre<-as.factor(eligible_subjects$study_centre)

```


### Get age of sample collection
```{r}
metadata <-read.csv("data/v1 2127B link to CHILD ID with age.csv")
metadata <- merge(metadata, alldata, by.x = "subjectnumber", by.y = "subject_id", all.x = TRUE, all.y = FALSE)
metadata <- merge(metadata, eligible_subjects, by.x = "subjectnumber", by.y = "subject_id", all.x = TRUE, all.y = FALSE)

```




### Merge the metadata dataframe with the data dataframe
```{r}
# Merge subject_id, visit, collect_age
merged_data <- merge(data, metadata, by.x = "SampleID", by.y = "SampleID", all.x = TRUE, all.y = FALSE)
```


### Systemic antibiotic use
```{r}
systemic_antibiotics<-read.csv("data/systemic_antibiotics_taken_after_hospital_discharge_to_1y.csv")

# Identify participants that took 3 or more systemic antibiotics before the age of 1
systemic_antibiotics$systemic_ab_gte_3 <- systemic_antibiotics$num_systemic_antibotics_birth_1y >=3 
systemic_antibiotics$systemic_ab_gte_3 <- as.integer(systemic_antibiotics$systemic_ab_gte_3)
systemic_antibiotics$systemic_ab_gte_3 <- as.factor(systemic_antibiotics$systemic_ab_gte_3)

# Identify participants that took 2 or more systemic antibiotics before the age of 1
systemic_antibiotics$systemic_ab_gte_2 <- systemic_antibiotics$num_systemic_antibotics_birth_1y >=2
systemic_antibiotics$systemic_ab_gte_2 <- as.integer(systemic_antibiotics$systemic_ab_gte_2)
systemic_antibiotics$systemic_ab_gte_2 <- as.factor(systemic_antibiotics$systemic_ab_gte_2)

merged_data <- merge(merged_data, systemic_antibiotics, by.x = "subjectnumber", by.y = "subject_id", all.x = TRUE, all.y = FALSE)     

```


### Updates the assay names so that they match the gene names used by clusterProfiler 
```{r}
# Fix some assay names so that they are compatible with the gsea done below

merged_data$Assay <- str_replace(merged_data$Assay, "-", "")
merged_data$Assay <- str_replace(merged_data$Assay, " ", "")
merged_data$Assay <- str_replace(merged_data$Assay, "alpha", "A")
merged_data$Assay <- str_replace(merged_data$Assay, "beta", "B")
merged_data$Assay <- str_replace(merged_data$Assay, "gamma", "G")
merged_data$Assay <- str_replace(merged_data$Assay, "MCP3", "CCL7")
merged_data$Assay <- str_replace(merged_data$Assay, "MCP2", "CCL8")
merged_data$Assay <- str_replace(merged_data$Assay, "Flt3L", "FLT3LG")
merged_data$Assay <- str_replace(merged_data$Assay, "BetaNGF", "NGF")
merged_data$Assay <- str_replace(merged_data$Assay, "ST1A1", "STP")
merged_data$Assay <- str_replace(merged_data$Assay, "NT3", "NTF3")
merged_data$Assay <- str_replace(merged_data$Assay, "OPG", "TNFRSF11B")
merged_data$Assay <- str_replace(merged_data$Assay, "SCP", "KITLG")
merged_data$Assay <- str_replace(merged_data$Assay, "LAPTGFB-1", "TGFB1")
merged_data$Assay <- str_replace(merged_data$Assay, "LAPTGFB1", "TGFB1")

merged_data$Assay <- str_replace(merged_data$Assay, "TWEAK", "TNFSF12")
merged_data$Assay <- str_replace(merged_data$Assay, "TRAIL", "TNFSF10")
merged_data$Assay <- str_replace(merged_data$Assay, "PDL1", "Q9NZQ7")
merged_data$Assay <- str_replace(merged_data$Assay, "TNFB", "LTA")
merged_data$Assay <- str_replace(merged_data$Assay, "TRANCE", "TNFSF11")
merged_data$Assay <- str_replace(merged_data$Assay, "MCP4", "CCL13")
merged_data$Assay <- str_replace(merged_data$Assay, "MCP1", "CCL2")
merged_data$Assay <- str_replace(merged_data$Assay, "IL8", "CXCL8")
merged_data$Assay <- str_replace(merged_data$Assay, "uPA", "PLAU")
merged_data$Assay <- str_replace(merged_data$Assay, "IL8", "CXCL8")

```


# Create a separate data frame for each olink dataset based on age
```{r}
olink_5y <- merged_data[merged_data$time == '5Y', ]
olink_1y <- merged_data[merged_data$time == '1Y', ]

```





# Associations between upper quartile adrenic acid and 1-year pathways

```{r}

olink_1y_filtered = olink_1y
olink_1y_filtered <- subset(olink_1y_filtered, !is.na(upper_quartile_adrenic_acid))

ttest_results_upper_quartile_adrenic_acid_1y <- olink_ttest(
  df = olink_1y_filtered,
  variable = "upper_quartile_adrenic_acid",
  alternative = "two.sided")

write.csv(ttest_results_upper_quartile_adrenic_acid_1y,"ttest_results_monthly_upper_quartile_adrenic_acid_1y.csv")
ttest_results_upper_quartile_adrenic_acid_1y[order(ttest_results_upper_quartile_adrenic_acid_1y$estimate, decreasing = FALSE),]
```



```{r}

set.seed(42) 

gsea_upper_quartile_adrenic_acid_1y_results <- olink_pathway_enrichment(data = olink_1y_filtered, ontology='MSigDb',test_results = ttest_results_upper_quartile_adrenic_acid_1y)
gsea_upper_quartile_adrenic_acid_1y_results$exposure = 'Upper quartile adrenic acid'

# Drop rows with 'p.adjust' less than 0.05
gsea_upper_quartile_adrenic_acid_1y_results <- gsea_upper_quartile_adrenic_acid_1y_results %>% filter(p.adjust < 0.05)
gsea_upper_quartile_adrenic_acid_1y_results <- gsea_upper_quartile_adrenic_acid_1y_results[c("exposure", setdiff(names(gsea_upper_quartile_adrenic_acid_1y_results), "exposure"))]
write.csv(gsea_upper_quartile_adrenic_acid_1y_results,"olink_1y_gsea_upper_quartile_adrenic_acid_1y.csv")
gsea_upper_quartile_adrenic_acid_1y_results
```

# Associations between upper quartile adrenic acid and 5-year pathways

```{r}
olink_5y_filtered <- subset(olink_5y, !is.na(upper_quartile_adrenic_acid))


ttest_results_monthly_upper_quartile_adrenic_acid_5y <- olink_ttest(
  df = olink_5y_filtered,
  variable = "upper_quartile_adrenic_acid",
  alternative = "two.sided")
write.csv(ttest_results_monthly_upper_quartile_adrenic_acid_5y,"ttest_results_monthly_upper_quartile_adrenic_acid_5y.csv")
ttest_results_monthly_upper_quartile_adrenic_acid_5y[order(ttest_results_monthly_upper_quartile_adrenic_acid_5y$estimate, decreasing = FALSE),]
```

```{r}
olink_5y_filtered |> 
  filter(upper_quartile_adrenic_acid == 0) |> 
  distinct(subjectnumber) |> 
  nrow()
```

```{r}

set.seed(42) 

gsea_results_upper_quartile_adrenic_acid_5y <- olink_pathway_enrichment(data = olink_5y_filtered, ontology='MSigDb',test_results = ttest_results_monthly_upper_quartile_adrenic_acid_5y)
gsea_results_upper_quartile_adrenic_acid_5y$exposure = 'Upper quartile adrenic acid'

# Drop rows with 'p.adjust' less than 0.05
gsea_results_upper_quartile_adrenic_acid_5y <- gsea_results_upper_quartile_adrenic_acid_5y %>% filter(p.adjust < 0.05)
gsea_results_upper_quartile_adrenic_acid_5y <- gsea_results_upper_quartile_adrenic_acid_5y[c("exposure", setdiff(names(gsea_results_upper_quartile_adrenic_acid_5y), "exposure"))]
write.csv(gsea_results_upper_quartile_adrenic_acid_5y,"olink_5y_upper_quartile_adrenic_acid_5y.csv")

gsea_results_upper_quartile_adrenic_acid_5y[order(gsea_results_upper_quartile_adrenic_acid_5y$NES, decreasing = FALSE),]
```


